\section{Discussion}
\label{sec:discussion}

The experimental results yield several key insights regarding the capabilities and limitations of Visual Foundation Models (VFMs) in dense semantic matching tasks.

\paragraph{Backbone Comparison: Semantic vs. Structural Awareness.}
Our evaluation highlights a clear distinction in how different backbones encode visual information. The DINO family (DINOv2 and DINOv3) demonstrates superior semantic abstraction, accurately linking object parts even across significant domain shifts (e.g., photos to paintings). Conversely, the Segment Anything Model (SAM) exhibits exceptionally strong structural and boundary awareness due to its mask-based pre-training. However, SAM's class-agnostic nature occasionally leads to geometrically precise but semantically incorrect matches compared to DINO.
\TODO{Add quantitative backbone comparison: report best PCK@0.10 per backbone at baseline, citing specific values from Table~\ref{tab:quantitative_results}. Mention which backbone achieves the highest zero-shot PCK and whether the ranking holds after fine-tuning.}

\paragraph{Impact of Inference Refinement.}
The transition from a standard discrete \textit{argmax} to the Windowed Soft-Argmax (WSA) systematically improved the Percentage of Correct Keypoints (PCK) across all precision thresholds. By enabling sub-pixel coordinate prediction, WSA effectively mitigates the discretization error inherent in patch-based ViT architectures and reduces the model's sensitivity to isolated noise spikes in the similarity maps.
\TODO{Quantify the average WSA gain ($\Delta$PCK@0.10) across all backbones and stages from \texttt{runs/paper\_figures/figures/wsa\_gain\_a0.1.pdf}. Discuss whether WSA gain is larger on the frozen baseline or on fine-tuned models.}

\paragraph{Efficiency of Low-Rank Adaptation.}
Full fine-tuning of large-scale ViT backbones is highly inefficient. The implementation of LoRA on the Multilayer Perceptron (MLP) layers of the final Transformer blocks proved to be a highly effective compromise: it significantly reduces the number of trainable parameters and memory footprint while providing sufficient capacity to adapt the frozen zero-shot features to the geometric nuances of the SPair-71k benchmark. Our LoRA depth sweep ($N \in \{1,2,4\}$) further characterizes how adaptation depth interacts with parameter efficiency, enabling a fine-grained accuracy–cost trade-off analysis. Measured inference throughput and peak GPU memory (reported in \texttt{efficiency\_results.json}) allow direct comparison of all configurations on the same hardware.
\TODO{Report exact trainable parameter counts (from AML\_Analysis.ipynb §H, \texttt{runs/paper\_figures/tables/efficiency.csv}): how many parameters does LoRA add vs.\ full fine-tuning of the last $N$ blocks? Report the PCK@0.10 gap between LoRA and the best FT configuration, and the throughput ratio between the most and least efficient configurations.}

\paragraph{Effect of Photometric Augmentation.}
The augmentation ablation (Section~\ref{sec:ablations}) isolates the contribution of photometric transformations applied during fine-tuning. Augmentation is expected to improve generalization particularly for backbones that are already strongly pretrained, where fine-tuning with clean data can lead to feature collapse on the training distribution.
\TODO{Quantify the average $\Delta$PCK@0.10 attributable to augmentation across all backbone $\times$ block configurations. Discuss whether the benefit is consistent or backbone-specific, referencing \texttt{runs/paper\_figures/figures/aug\_ablation.pdf}.}

\paragraph{Limitations and Failure Cases.}
Despite the improvements, semantic correspondence remains challenging under extreme conditions. Qualitative analysis reveals that the models still struggle with severe truncations and symmetry ambiguity (e.g., consistently distinguishing a left eye from a right eye when the object undergoes a severe out-of-plane rotation). Addressing these geometry-aware semantic failures remains an open challenge for future work.
\TODO{Add per-difficulty breakdown (viewpoint / scale / truncation / occlusion) from AML\_Analysis.ipynb §E (\texttt{runs/paper\_figures/tables/per\_difficulty\_a0.1.csv}). Identify which difficulty factor causes the largest PCK drop. Add per-category heatmap figure from §D if space allows.}

\paragraph{Training Protocol Constraints and Omitted Optimisations.}
Several standard training improvements were deliberately excluded from our protocol in order to preserve the comparability of results across all backbone and adaptation configurations.

\textbf{Hyperparameter search strategy.} The only hyperparameter subjected to a systematic sweep in this work is the adaptation depth $N \in \{1, 2, 4\}$, which constitutes an exhaustive \textit{grid search} over a small, manually predefined grid. All remaining hyperparameters (learning rate, weight decay, LoRA rank and $\alpha$, batch size, early-stopping patience) were fixed \textit{a priori} based on standard practice and project specification, without any automated search. More principled strategies exist along a spectrum of cost and effectiveness: \textit{random search} samples configurations from predefined distributions and often outperforms grid search in high-dimensional spaces; \textit{coarse-to-fine} search first identifies promising regions with short training runs and then refines within them; \textit{Bayesian optimisation} builds a probabilistic surrogate of the objective function (e.g.\ via Gaussian Processes or tree-based models) and uses it to select the most informative next configuration, achieving good results with far fewer evaluations than grid or random search. All of these approaches were excluded for the same reason: each evaluation requires training the full pipeline to convergence, which on our hardware amounts to several days per complete sweep. Applying any of these methods to even a small hyperparameter space would therefore be computationally intractable within the project timeline.

\textbf{Learning rate scheduling.} Techniques such as cosine annealing with warm restarts or adaptive decay (e.g.\ \textit{ReduceLROnPlateau}) could reduce the oscillatory behaviour observed in the validation loss during fine-tuning and potentially allow convergence to lower minima in fewer epochs. \textit{ReduceLROnPlateau}, in particular, would interact naturally with our early-stopping criterion. However, introducing a scheduler adds further hyperparameters (cycle length, minimum LR, decay factor) whose tuning cost would itself require multiple full training runs per backbone, undermining the controlled comparison that is the primary objective of this study.

\textbf{Early-stopping patience.} A patience of 7 epochs was fixed uniformly across all 18 training stages. Observed validation loss curves exhibit non-monotonic oscillations of up to $\pm0.05$, and in several runs new minima were reached after 5--6 consecutive non-improving epochs, suggesting that a moderately higher patience (e.g.\ 10--12) could occasionally recover better checkpoints. Nevertheless, varying this value per backbone or per adaptation depth would invalidate the cross-configuration comparison that is central to our ablation methodology.

Crucially, the computational cost of a full re-run is prohibitive in the context of this project. Even a single backbone sweep (3 fine-tuning stages and 3 LoRA stages) requires approximately 1--2 days of continuous computation on the hardware described in Section~\ref{sec:experiments}. A complete re-run of all 18 training stages to evaluate a single hyperparameter variation would therefore require on the order of one week of uninterrupted GPU time. For this reason, all optimisation decisions were fixed \textit{a priori} and held constant throughout the entire experimental campaign, ensuring full reproducibility and fair comparison at the cost of potentially sub-optimal individual configurations.